---
title: "Comparison of Algorithms for Estimating Latent Class Models"
description: |
  My master's thesis!
author:
  - name: Jessica
categories:
  - research
date: 09-06-2025
output:
  distill::distill_article:
    self_contained: false
---

```{r, include = F}
knitr::opts_chunk$set(echo =F)
library(kableExtra)
library(tidyverse)
```

In august I received my master's degree in Psychology! It took me a long time to get here and today I will be summarizing the work I did for my master's thesis which is titled: Comparison of Algorithms for Estimating Latent Class Models. There was a lot of trial and error, code that had to be rerun many times and rewriting a lot of results, but I'm proud of the work I've done. More importantly, I have a lot more ideas to follow-up on as a result of what I found in this thesis. For brevity I'm going to assume familiarity of latent class models.

### Motivation
In Maximum Likelihood Estimation (MLE) - a framework of obtaining parameter estimates that I'm focusing on- the goal is end up with parameters that maximize the probability of observing our data, the response patterns that we collected from respondents. 

Under MLE, latent class models are commonly estimated using the Expectation-Maximization algorithm (EM). EM has two steps: the first involves computing the expected value of the complete data likelihood - which in the case of latent class models is the posterior probability across all classes and individuals. These values are calculated based on the current parameter estimates- the first iteration of EM will be calculated based on randomly generated starting values.

In the Maximization (M-step), these posterior probabilities are then summed and averaged to get new parameter estimates for both item response probabilities and class prevalences. Then we use these new parameter estimates to calculate new posterior probabilities (E-step) and subsequently updating the parameter estimates again. This process continues until there is either minimal changes in the likelihood value or parameter estimates after an iteration.

There are some great theoretical properties of the EM algorithm, one of them being that as the EM algorithm progresses, the likelihood will never decrease. This is important as our goal is to maximize the likelihood, so we should always be searching for parameter values that lead to better a likelihood.

A common gripe about the EM algorithm is that the steps it takes towards the optimal set of parameter estimates can be extremely small, leading to slow convergence rates. This has led researchers to find ways to speed up the EM algorithm and explore other algorithms.

As I was exploring estimation algorithms, derivative based optimization caught my eye. Mainly because at the time I understood the intuition behind using first and second derivatives to inform next steps in estimation more than the steps that EM was taking. First and second derivatives let us know in what direction the function is increasing/decreasing, and we can use that information to only move in a direction that we desire. Upon doing further research I learned that calculating second derivatives can be computationally intense, although once calculated, will result in larger steps being taken at each step of the estimation algorithm. The stark difference in how these algorithms moved intrigued me. From my personal experience reading articles using latent class models, there is not much discussion surrounding the choice of estimation algorithms. Rather, whatever is the default algorithm of the software used is accepted. 

Without understanding much about estimation at the time, I hypothesized that it was possible that different estimation algorithms could lead to different parameter estimates for the same model and dataset, since they relied on different quantities and rules for parameter searching. I didn't really care if my hypothesis was right or wrong, but I couldn't find any literature about my curiosities! This is what motivated me to design my first every simulation study so at least I could begin to scratch the surface of the question.

One additional aspect of estimation that I was curious about is local solutions. Normally when estimating latent class models it is strongly recommended to use multiple sets of starting values. This is because the complexity of the parameter space - with latent class prevalence parameters that have to add up to 1, small sample sizes, etc- could lead a set of badly located starting values to an area of the parameter space that is a maximum or peak in its tiny neighborhood, but overall is not the maximum likelihood solution across the whole space. By using a variety of starting values that are spread out well throughout the parameter space, the largest maximum likelihood value we find across these multiple starts has a higher chance (BUT is not guaranteed) to be the global maximum. Typically if we initialize our estimation algorithm with hundreds of starting values, we will keep the solution with the largest maximum likelihood and discard the other iterations. I am interested in whether the presence/frequency/location of local solutions has an impact on estimation quality. In this specific case I define estimation quality as bias- distance of the estimated parameters to the true parameters. Other outcomes of interest may include class enumeration, which is choosing the correct number of classes to describe the dataset.

When we simulate data we know the true number of classes, their size, and the actual parameter values. When using empirical data we know none of this and can only rely on limited known information such as the number of items or sample size for any indication of what quality of parameter estimates we might get. If we can incorporate information about local solutions- perhaps their frequency, or if they're occurring very close to the maximum solution- we can use this information as another piece in the puzzle to understanding how well our model is being estimated.

Overall I have two major research questions:
1. Are there differences in how NR and EM navigate the parameter space and find the optimal solution? 
2. Can examining local solutions such as the second best likelihood solution or the most common likelihood solution provide us with more information about the bias of our parameter estimates?

A natural extension to my existing research questions is whether NR and EM differ in the local solutions found, and whether local solutions from either or both algorithms are useful for assessing the quality of our solution.

Although my set of specific research questions have not been well researched, I will share previous studies regarding local solutions and comparison of estimation algorithms. As the home page of google scholar quotes- "stand on the shoulder of giants".  Research is iterative and builds on pre-existing work, so it is important to highlight what has already been done, and to use that to inform future work.

### Previous Research

There are two papers I want to highlight that inspired my simulation design have helped me shape my ideas and thought process.

The first paper is by [Chen et al, 2021](https://www.tandfonline.com/doi/full/10.1080/03610918.2020.1764034) - where the authors compared the performance of 2 derivative based algorithms against EM for estimation of latent class models. They found that the derivative based algorithms converged to the same likelihood solution as EM, and did it in fewer steps. You would think that this study answered my curiosities, but I felt that their study design and method still left a few gaps. The simulated datasets they used were limited in scope and in my opinion, not reflective of real datasets researchers use latent class models on. They also did not provide detailed information about start value generation or frequency of solutions found. Their paper is the first to conduct a simulation study comparing estimation algorithms in latent class, so these lack of details are not critiques against the authors! These are gaps in the research that I hope to fill. 

Another paper by [Shireman et al. 2016](https://pubmed.ncbi.nlm.nih.gov/27494191/) serves as the inspiration for my interest in local solutions. The authors conducted a simulation study investigating whether the number of local solutions produced by fitting mixture models was associated with bias and class enumeration. They found a moderate negative relationship (correlation of -0.50) between classification accuracy and number of local solutions, but no correlation between estimation bias and the number of local solutions. The authors suggest using local optima to make “rough inferences” about the quality of cluster recovery, but do not provide specific recommendations or further guidelines. I wanted to conduct my own simulation study similar to this paper, but using latent class models specifically instead.

### simulation design and process

In order to conduct a simulation study that is generalizable and useful to applied researchers, it should cover many different factors and data characteristics. In the interest of time (and finishing my masters thesis), I narrowed down my study to include only 8 simulation conditions, although I wanted to do much more!  I varied sample size (400 vs 1000), class prevalence balance (equally sized classes vs not equally sized), and amount of measurement error (low vs high). Measurement error refers to the item response probabilities. If theoretically individuals in a class should all be endorsing item 1, an item response probability of .95 is considered low because there is only a 5% chance someone will not endorse item 1. If an item response probability is .65, that means theres a 35% chance someone will not endorse item 1 even when they are supposed to. We don't know whether our item response probabilities have low or high measurement error in our empirical data, but we know in our simulated data.

These three factors were fully crossed to create 8 simulation conditions. For each condition I simulated 10 datasets and used 500 starting values for each dataset. 

As for the estimation algorithms, I decided to compare Newton-Raphson and EM.
Many different derivative based optimization methods exist, but for the sake of this project I narrowed my focus to NR. The algorithm is relatively intuitive and estimating latent class models using NR has already been programmed in STATA's gsem package. I tried to program NR on my own in R but was not successfully able to do so. For EM, I used code written in R that mimics the popular poLCA package to estimate the models.

Using two different software to do a comparison study is definitely questionable and vulnerable to criticism, and I acknowledge that. I hope to be able to program NR or another derivative based algorithm in R in the near future. 

Both NR and EM were given the same sets of starting values for each dataset and simulation condition.

### Results

I will highlight 3 main results from my thesis.

1. In most simulation conditions, NR and EM converged to the same maximum likelihood solution overall, although there were a few differences in the worst simulation condition (small sample size, high measurement error, and unbalanced class prevalences). 
```{r}
rho_table3 <- tribble(
  ~TrueParameters,     ~Class1, ~Class2, ~Class3, ~Class4,
  "Class Prevalence", 0.70,    0.15, 0.10, 0.05,
  "Item 1",  0.65,    0.65,    0.35,    0.35,
  "Item 2",  0.65,    0.65,    0.35,    0.35,
  "Item 3",  0.65,    0.65,    0.35,    0.35,
  "Item 4",  0.65,    0.65,    0.35,    0.35,
  "Item 5",  0.65,    0.65,    0.35,    0.35,
  "Item 6",  0.65,    0.65,    0.65,    0.35,
  "Item 7",  0.65,    0.65,    0.65,    0.35,
  "Item 8",  0.65,    0.65,    0.65,    0.35,
  "Item 9",  0.65,    0.65,    0.65,    0.35,
  "Item 10", 0.65,    0.65,    0.65,    0.35
)

# EM 
rho_table1 <- tribble(
  ~EM,      ~Class1, ~Class2, ~Class3, ~Class4,
  "Class Prevalence", 0.82,    0.09, 0.06, 0.03,
  "Item 1",  0.65,    0.72,    0.51,    0.00,
  "Item 2",  0.66,    0.56,    0.33,    0.21,
  "Item 3",  0.62,    0.54,    0.28,    0.52,
  "Item 4",  0.65,    0.56,    0.01,    0.54,
  "Item 5",  0.65,    0.71,    0.14,    0.27,
  "Item 6",  0.61,    0.34,    0.73,    0.19,
  "Item 7",  0.63,    0.08,    1.00,    0.51,
  "Item 8",  0.65,    0.35,    0.72,    0.00,
  "Item 9",  0.68,    0.12,    0.33,    0.68,
  "Item 10", 0.65,    0.23,    0.60,    0.16
)

# NR 
rho_table2 <- tribble(
  ~NR,      ~Class1, ~Class2, ~Class3, ~Class4,
 "Class Prevalence", 0.60,    0.28, 0.04, 0.09,
  "Item 1",  0.69,    0.58,    0.30,    0.53,
  "Item 2",  0.70,    0.54,    0.44,    0.35,
  "Item 3",  0.61,    0.56,    1.00,    0.42,
  "Item 4",  0.66,    0.56,    1.00,    0.20,
  "Item 5",  0.74,    0.58,    0.06,    0.09,
  "Item 6",  0.60,    0.43,    1.00,    0.74,
  "Item 7",  0.64,    0.37,    0.53,    1.00,
  "Item 8",  0.69,    0.36,    0.70,    0.71,
  "Item 9",  0.67,    0.45,    1.00,    0.51,
  "Item 10", 0.72,    0.30,    0.60,    0.66
)


# differene between EM - NR

rho_diff <- tribble(
  ~Differences, ~Class1, ~Class2, ~Class3, ~Class4,
 "Class Prevalence", 0.22,    -0.19, 0.02, -0.06,
  "Item 1",     -0.04,    0.14,     0.21,   -0.53,
  "Item 2",     -0.04,    0.02,    -0.11,   -0.14,
  "Item 3",      0.01,   -0.02,    -0.72,    0.10,
  "Item 4",     -0.01,    0.00,    -0.99,    0.34,
  "Item 5",     -0.09,    0.13,     0.08,    0.18,
  "Item 6",      0.01,   -0.09,    -0.27,   -0.55,
  "Item 7",     -0.01,   -0.29,     0.47,   -0.49,
  "Item 8",     -0.04,   -0.01,     0.02,   -0.71,
  "Item 9",      0.01,   -0.33,    -0.67,    0.17,
  "Item 10",    -0.07,   -0.07,     0.00,   -0.50
)
kable(rho_table3, caption = "True Data Generating Parameters",booktabs = T)
```

```{r}
kable(rho_table1, caption = "EM parameter estimates",booktabs = T)
```

```{r}
kable(rho_table2, caption = "NR parameter estimates",booktabs = T)
```

```{r}
kable(rho_diff, caption = "Differences",booktabs = T)
```

The tables above show the rho parameter estimates for simulation 3 dataset 1 for EM, NR, and differences between the two. In this dataset the maximum LL solution from NR was higher than EM. For class 1 and 2 of both EM and NR solutions, our interpretation and labeling of the class would be virtually the same: in this class respondents are likely to respond positively to all items. Although in class 2 there are more distinct differences (differences in probabilities greater than .1) between parameter estimates produced by the two algorithms, the interpretation remains the same. However, for classes 3 and 4 there are larger differences in the final solution. In class 3, the probability of positive endorsement for item 9 is .33 for EM, while it is 1.0 in NR. According to the EM solution someone who is a member of class 3 only has a 33% chance of saying yes to item 9, while in the NR solution they have a 100% chance of saying yes. Lastly, in EM class 4 items 3 and 4 would be interpreted as positive endorsement while the same parameter estimates from NR would be interpreted as no endorsement in NR. Examining the bottom table shows large differences in parameter estimates between EM and NR for classes 3 and 4. These are substantial differences that would impact our overall interpretation of the latent class solution. In both NR and EM solutions boundary estimates were observed, indicating potential instability or model overparameterization. As such, substantive interpretation of the final solution should be made cautiously, and alternative class solutions or model specifications may need to be considered. Lastly, while in this case we are focusing on differences in the maximum LL solution from the two algorithms, it should be noted that both solutions deviate from the true data generating parameters substantially, highlighting the challenges of parameter recovery in less than ideal conditions. These differences are important to highlight because this means that algorithm choice matters. If we get different answers depending on algorithm choice, this has a serious consequence on validity.   

Furthermore, these results pertain only to overall highest likelihood solution. There are actually many differences between what NR and EM converge to when looking on at an iteration by iteration basis. 

```{r}
individual <- readRDS("individual_new.Rds")
individual <- individual %>%
  mutate(diff = NR - EM,
         equal = ifelse(round(NR,3) == round(EM,3) ,T,F))

temp9 <- individual %>%
  filter(NR_converged ==1 )%>%
  group_by(sim_number) %>%
  summarize( 
    eq = sum(equal == T ,na.rm = T)/n(),
          maxmaxeq = sum(NRmax == T & EMmax == T & equal == T,na.rm = T)/n(),
          nonoeq = sum(NRmax == F & EMmax == F & equal == T,na.rm = T)/n(),
          maxnoeq = sum(NRmax == T & EMmax == F & equal == T,na.rm = T)/n(),
     nomaxeq = sum(NRmax == F & EMmax == T & equal == T,na.rm = T)/n())

temp9 <- temp9 %>% mutate(across(where(is.numeric), ~ round(.x, 2)))
row.names(temp9) <- 1:nrow(temp9)
colnames(temp9) <- c("Simulation Number","Proportion of Starting Values Where NR and EM LL are Equal",
                     "NR and EM LL are Equal and are both Max LL",
                     "NR and EM LL are Equal and are both NOT Max LL",
                     "NR and EM LL are Equal: NR is Max and EM is not",
                     "NR and EM LL are Equal: EM is Max and NR is not")

kbl(temp9[,-c(5,6)], align = "c", caption = "Proportion of Starting Values Where NR and EM Converge to Equal LL") %>%
  column_spec(1:2, bold = T) %>%
  collapse_rows(columns = 1:2, valign = "top") %>%
  kable_paper(full_width = F)   %>%  kable_styling(latex_options =  c("scale_down","striped"))
```

The table above shows the rates of agreement between the two algorithms across conditions. In the worst simulation condition (referred to as simulation 4), only 40% of starting values had NR and EM converging to the same value. There aren't any consistent trends that show that NR or EM converge to higher likelihood solutions more often, but their lack of agreement is shocking.

2.  Examining the second best likelihood solutions and most common likelihood solution (if it is not the maximum) could potentially be useful. I found that in the worst simulation condition, there was one instance where the second best solution was extremely close in likelihood value (to the thousandths decimal place), but produced a completely different solution. This might be an indication of unstable parameter estimates! In the worst simulation condition, which also had the highest bias, the most common solution was very often not the maximum likelihood solution. Another interesting insight is that the most common solution was occasionally closer to the true parameter values than the maximum in poor data conditions. However, I did not find a consistent trend of this occurring in specific conditions. 

3. The number of local solutions is positively associated with bias, but does not explain much of the variation in bias. Based on the results of this simulation study, there are no strong conclusions we can draw about the bias of the solution based on the number of local solutions. These results are based on how I model the relationship between the two, and there could be issues with my methodology which I will discuss in the last section.


### Can of Worms

My research is significant in that it challenges implicit assumptions that using different estimation algorithms on the same data set to estimate a latent class model will get you the same solution. Even if the final maximum solution between NR and EM only differs in rare occasions when data characteristics are poor, that is STILL worth highlighting. I believe that the empirical data we have is closer to poor data characteristics we simulate rather than ideal ones. It also opens the conversation about how we can use information about local solutions to identify estimation issues. My research does not suggest anything conclusive like X number of local solutions means that the final solution will be far away from the truth. The conclusions I can make are quite weak, similar to those made in Shireman et al. (2016). However, I don't think this means that local solutions are useless and that we shouldn't continue exploring. Rather, I want to spend more time thinking about how to model the relationship between local solutions and bias more carefully. How should we quantify local solutions, and how should we quantify bias?

I have three issues that I need to address in my future work that pertain to those two research questions. One, is start value generation. How I decide to generate start values is directly connected to how many local solutions will be found in the estimation process. In this simulation study I chose to generate start values by randomly assigning observations to classes and obtaining class prevalence and item response probabilities from this initial assignment. This method results in fewer boundary estimates as starting values, but may not ensure sufficient coverage of the parameter space. Before I conduct any further research I should look into various start value generation methods and think carefully about which one to use.  All of my results from this study are contingent on the generation method that I used.  

Second is the issue of label switching. This occurs when class 1 and class 2 might switch labels but still have the same likelihood value. This can make computing the bias challenging because I might not have matched the parameters of the estimated classes to the parameters of the true classes correctly. I am still trying to figure out how to ensure accuracy, and am planning to look into Bayesian markov models which have addressed this issue.

My last issue is that I'm not 100% sure if I'm modeling the relationship between the # of local solutions and bias in the most meaningful way. In my thesis I fit a multilevel model predicting bias as a function of sample size, class prevalence balance, measurement error, and number of local solutions, ultimately finding that the number of local solutions wasn't a useful predictor. However, with empirical data we wouldn't know the true measurement error or class prevalence balance, so should I fit a model with fewer predictors? Also, is the coefficient I get from this model meaningful? What does a .5 increase in bias, for example, really mean in terms of the parameter estimates and the story the classes tell? Before I get really wrapped up in bias as a metric, it is worth taking the time to really explore what "ideal" values of bias are versus "poor" values, if those exist.

Another question that I just thought of is: is it even meaningful to look for local solutions with closer distance to the true parameters? maximum likelihood estimation is about maximizing likelihood and not recovering parameters, so using the results produced from a MLE framework and fixating on solutions that are not the MLE seems a bit off to me. I definitely need to do some research to 

I have a lot to chew on in terms of my next steps, but I'm super excited! Writing out all the pitfalls and unknowns of my research is actually really motivating. As the new school year begins, I hope that I can make good progress and contribute something useful to the quantitative methodology field!