---
title: "Another exploration of local solutions and bias "
description: |
  The quest continues
author:
  - name: Jessica
categories:
  - research
date: 10-03-2025
output:
  distill::distill_article:
    self_contained: false
---

```{r}
library(lmerTest)
library(tidyverse)
fin <- readRDS("fin.Rds")

fin <- fin %>%
  mutate(distance =  max_val - secondLL)

fin <- fin %>%
  mutate(distance = ifelse(distance < 0,0,distance))
```

# ARI vs Classification Accuracy

CA and ARI are highly correlated when measurement error is low, but once measurement error is high ARI is a more conservative indicator. Based on the formulation of ARI this makes sense, as ARI accounts for the fact that individuals could be correctly classified due to chance. Thus, I will use ARI as the classification accuracy metric. 
```{r}
#cor(fin$m_ARI,fin$m_CA)
ggplot(data = fin, aes(x = m_ARI, y = m_CA, color = me)) + geom_point()
```

# Relationship between local solutions, bias, and ARI

Based on the pairs plot, we can identify a few trends between the 3 variables. Percentage of unique solutions and bias are positively correlated, and it appears that there is a ceiling value of around .25 for bias. 

ARI is bimodal, and there doesn't seem to be a clear relationship between percentage of unique solutions and ARI for either mode. It's possible to have low uniq_percentage and perfect or or super low ARI. 

```{r}
pairs(fin[,c(3,9,15)])
```

ARI is bimodal ---split on extent of measurement error

When measurement error is low: high ARI almost guarantees low bias.., but just because ARI is low doesn't mean that bias is going to be high...

When measurement error is high: even if ARI is low, like in the .2 range ..its still possible for bias to be low..
```{r}

ggplot(data = fin, aes(x = m_ARI, y = bias, color = me))+ geom_point()

#ggplot(data = fin, aes(x = m_ARI, y = bias, color = nitem))+ geom_point()
```

# Simulation factors predicting bias and ARI 

Much like what I found during my master's thesis, measurement error explains most of the variation in bias. The number of classes also explains a lot too. 
```{r}
mod <- lmer(bias~ uniq + ss + me + prev + nclass + nitem + (1 | sim_number), data = fin)
#summary(mod)
#car::Anova(mod, type = 3) 
effectsize::eta_squared(mod)
```

```{r}
# using only things we could know a priori
mod <- lmer(bias~ uniq + ss +nitem + (1 | sim_number), data = fin)
summary(mod)
car::Anova(mod, type = 3) 
effectsize::eta_squared(mod)

# is this useful? it's not like we know sim_number IRL
mod2 <- lmer(bias~ uniq + ss +nitem + (1 +uniq | sim_number), data = fin)
anova(mod,mod2)
```



```{r}
mod <- lmer(m_ARI~ uniq + ss + me + prev + nclass + nitem + (1 | sim_number), data = fin)
summary(mod)
car::Anova(mod, type = 3) 
effectsize::eta_squared(mod)
```

Are these bad rule of thumbs?

At 10% unique solutions - meaning about 50 different likelihood values - you're more likely to have item parameters with high measurement error... high bias, poor classification... regardless of sample size and number of items

Having a larger sample size increases the changes that you have items with high measurement error IF THE PERCENTAGE OF UNIQUE SOLUTIONS IS HIGH. 
```{r}

 ggplot(data = fin, aes(x = uniq_perc, y = bias, color = me)) + facet_wrap(~ nitem + ss)+ geom_point()
 
ggplot(data = fin, aes(x = uniq_perc, y = m_ARI, color = me)) + facet_wrap(~ nitem + ss)+ geom_point()

ggplot(data = fin, aes(x = uniq_perc, y = m_ARI, color = nclass)) + geom_point()
```

The correlation between percentage of unique solutions and bias is the strongest at .77 when 

Compared to shireman et al who found 0 correlation between MSPE and proportion of unique solutions...... we have much higher! Shireman et al is a mixture model paper not a latent class one. 
```{r}
temp <- fin %>% filter(ss == "small",nitem == 7)
cor(temp$bias,temp$uniq)

temp <- fin %>% filter(ss == "large",nitem == 7)
cor(temp$bias,temp$uniq)

temp <- fin %>% filter(ss == "small",nitem == 10)
cor(temp$bias,temp$uniq)

temp <- fin %>% filter(ss == "large",nitem == 10)
cor(temp$bias,temp$uniq)

```

```{r}
mod1 <- lmer(bias~ max_n + ss  + me +  prev + nclass +nitem + (1 | factor(sim_number)), data = fin)
summary(mod1)
car::Anova(mod1, type = 3) 
effectsize::eta_squared(mod1)

```



```{r}

ggplot(data = fin, aes(x = uniq_perc, y = m_ARI, color = bias)) + facet_wrap(~ nitem + ss)+ geom_point()

temp <- fin %>% filter(ss == "small",nitem == 7)
cor(temp$m_ARI,temp$uniq_perc)

temp <- fin %>% filter(ss == "large",nitem == 7)
cor(temp$m_ARI,temp$uniq_perc)

temp <- fin %>% filter(ss == "small",nitem == 10)
cor(temp$m_ARI,temp$uniq_perc)

temp <- fin %>% filter(ss == "large",nitem == 10)
cor(temp$m_ARI,temp$uniq_perc)

```


```{r}



temp <- fin %>% filter(ss == "small",nitem == 7)
cor(temp$m_ARI,temp$bias)

temp <- fin %>% filter(ss == "large",nitem == 7)
cor(temp$m_ARI,temp$bias)

temp <- fin %>% filter(ss == "small",nitem == 10)
cor(temp$m_ARI,temp$bias)

temp <- fin %>% filter(ss == "large",nitem == 10)
cor(temp$m_ARI,temp$bias)

```

if more than 10% of iterations converge to unique... you're prob gonna have high bias
BUT just because you have low unique_perc DOESNT automatically mean your solution is good
```{r}

ggplot(data = fin, aes(x = uniq_perc, y = bias, color = me)) + facet_wrap(~ nitem + ss)+ geom_point()
ggplot(data = fin, aes(x = uniq_perc, y = bias, color = me))+ geom_point()

```

# second LL analysis


relationship between distance to second LL and bias?

Just because the distance is 0 or small doesn't suggest anything.. BUT if the distance of the second best LL is really big... probably a good sign
```{r}
ggplot(data = fin, aes(x = distance, y = bias, color = secondLL_n))+ geom_point()
```

Not sure what the point of this would be.
```{r}
mod1 <- lmer(distance~  ss  + me +  prev + nclass +nitem + (1 | sim_number), data = temp)
summary(mod1)
car::Anova(mod1, type = 3) 
effectsize::eta_squared(mod1)

```

```{r}
ggplot(data = fin, aes(x = distance, y = bias, color = me))+ facet_wrap(~ nitem + ss)+ geom_point()

ggplot(data = fin, aes(x = secondLL_n, y = bias, color = me))+ facet_wrap(~ nitem + ss)+ geom_point()
```

